\makeatletter\@openrightfalse \chapter*{Warranty and intended usage}
\@openrighttrue\makeatother

\thispagestyle{empty}

The warranty on material and assembly is 6 months. If the freezer is defective,
it may be replaced by one that is functioning. In case of hardware
incompatibility, the price of the freezer will be refunded, after the part has
been returned, free of postage. This warranty only covers the hardware. The risk
for error free operation of the software and the application for specific tasks rest solely with the buyer or
user. There are no rights, directly or implied, of whatever form, concerning
software, exchange, improvements, refunds (fully or partially), changes, updating
etc. Warranty does not cover damage caused by transport, incorrect installation
or use, static discharges, pollution, attempted extensions and any alterations to
the electronics outside our workshop. There's also no warranty on any parts that
were not installed by us.

Usage of the \frz must not in any way infringe rights of third parties or
violate any laws or regulations. Any responsibility or claims by third parties
of any form, illegal use, damage to property or dead or living matter, and any
other kind of claim is not accepted by us in any way and are the sole
responsibility of the user. It's the sole responsibility of the user, owner or
buyer to prevent illegal use or any other damage that may arise from using the
product. Adults are responsible for their children. The user should be aware of
regulations or laws concerning the usage and seek council in case of doubt.
Claiming ignorance or lack of knowledge is not an option out.

The use of commercial names, trademarks, labels or manufacturer makes is not
done with the assumption that any of these items should be freely usable and no
rights of the items are intentionally claimed or assumed.
All references do not take any patents or other rights into account.
Any claim from manufacturers is strictly excluded. Use at your own risk.

This manual is protected by copyright and subject to Creative Commons license\linebreak
BY-NA-SA 3.0.\newline
\url{http://creativecommons.org/licenses/by-nc-sa/3.0/legalcode}. 



